\pkg{partykit}: A Toolkit for Recursive Partytioning

### 2011-09-07 pkg/partykit/inst/doc/partykit.Rnw
further improvements to vignette
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@220 255f4249-2263-4417-b1ea-8498779e27a7

\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
   \And Achim Zeileis\\Universit\"at \\ Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

@@

  The purpose of the package is to provide a toolkit with infrastructure for representing,
  summarizing, and visualizing tree-structured regression and classification models.
  Thus, the focus is not on \emph{inferring} such a tree structure from data but
  to \emph{represent} a given tree so that print/plotting and computing predictions
  can be performed in a standardized way. In particular, this unified infrastructure can be
  used for reading/coercing tree models from different sources
  (\pkg{rpart}, \pkg{RWeka}, PMML) yielding objects that share functionality for
  \code{print}/\code{plot}/\code{predict} methods.
}
\Keywords{recursive partitioning, regression trees, classification trees, decision trees}

@@

\section{Introduction}

@@

Hence, in a third step the recursive tree structure stored
in \code{pn} is coupled with the corresponding data in a \class{party}
object.
<<weather-party>>=
py <- party(pn, WeatherPlay)
py
@
In addition to the \class{partynode} and the \class{data.frame},
the function \fct{party} takes several further arguments including
\code{fitted} values, a \code{terms} object, arbitrary additional
\code{info}, and some further arguments (\code{...}):
\begin{Code}
  party(node, data, fitted = NULL, terms = NULL, ..., info = NULL)
\end{Code}

The tree \code{py} can then be easily visualized by
<<weahter-plot, eval=FALSE>>=
plot(py)
@
yielding Figure~\ref{weather-plot}.

\subsubsection*{More}

<<weather-party-print>>=
print(py,
  terminal_panel = function(node) paste(": play=", node$info, sep = ""))
@

%foo_node
%foo_split

%kidids_node(partynode_object)
%kidids_split(partynode_object)

\section{Splits}

A split is basically a function that maps data, 
more specifically a partitioning variable, to daugther nodes. 
Objects of class \code{partysplit} are designed to represent 
such functions and are set-up by the \code{partysplit()} 
constructor:
<<partysplit-1, echo = TRUE>>=
## binary split in numeric variable `Sepal.Length'
sl5 <- partysplit(which(names(iris) == "Sepal.Length"), breaks = 5)
class(sl5)
@
The internal structure of class \code{partysplit} contains
information about the partitioning variable, the split-points,
the handling of split-points, the treatment of observations with
missing values and the daughter nodes to send observations to:
<<partysplit-2, echo = TRUE>>=
unclass(sl5)
@
Here, the split is defined in the first variable (corresponds to
\code{Sepal.Length} in data frame \code{iris}) and the splitting rule
is \code{Sepal.Length} $\le 5$:
<<partysplit-3, echo = TRUE>>=
character_split(sl5, data = iris)
@
This representation of splits is completely abstract and, most importantly,
independent of any data. Now, data comes into play when we actually
want to perform splits:
<<partysplit-4, echo = TRUE>>= 
kidids_split(sl5, data = iris)
@
For each observation in \code{iris} the split is performed and the 
number of the daughter node to send this observation to
is returned. Of course, this is a very complicated way of saying
<<partysplit-5, echo = TRUE>>=
(!with(iris, Sepal.Length <= 5)) + 1
@

@@

Let $\mu = (-\infty, \mu_1, \dots, \mu_{l - 1}, \infty)$ denote the split points
($(\mu_1, \dots, \mu_{l - 1})$ = \code{break}). We are interested to split according to the 
information contained in the $i$th element of $x$ ($i$ = \code{varid}).
For numeric $x_i$, the split points are also numeric. If $x_i$ is a factor at
levels $1, \dots, K$, the default split points are $\mu = (-\infty, 1, \dots, K - 1, \infty)$.

@@

The slots of \code{partysplit} objects shall be accessed by the corresponding
accessor functions.

\section{Nodes}

Inner and terminal nodes are represented by objects of class \code{partynode}.
Each node has a unique identifier \code{id}. A node consisting only
of such an identifier (and possibly additional information  
in \code{info}) is a terminal node:
<<partynode-1, echo = TRUE>>=
n1 <- partynode(id = 1L)
is.terminal(n1)
print(n1)
@
Inner nodes have to have a primary split \code{split} and at least two
daugther nodes. The daugther nodes are objects of class \code{partynode}
itself and thus represent the recursive nature of this data structure.
The daugther nodes are pooled in a list \code{kids}. 
In addition, a list of \code{partysplit} objects offering 
surrogate splits can be supplied; a list of \code{partysplit} objects
in slot \code{surrogates} defines such additional splits (mostly used 
for handling missing values). Note that \code{partynode} objects aren't 
connected to the actual data.

@@

A number of methods are defined for \code{partynode} objects.
\code{is.partynode} checks if the argument is a valid \code{partynode}
object. \code{is.terminal} is \code{TRUE} for terminal nodes
and \code{FALSE} for inner nodes. The subset methods
return the \code{partynode} object corresponding to the \code{i}th
kid:
<<partynode-5, echo = TRUE>>=
n1[2]
@

The \code{as.partynode} and \code{as.list} methods can be used
to convert flat list structures into recursive \code{partynode}
objects and vice versa. \code{as.partynode} applied to
\code{partynode} objects renumbers the recursive nodes
starting with root node identifier \code{from}.

\code{length} gives the number of kid nodes of the root node,
\code{depth} the depth of the tree and \code{width}
the number of terminal nodes.

\section{Trees}

@@

Although tree structures can be represented by \code{partynode} objects,
a tree is more than a number of nodes and splits. More
information about (parts of the) corresponding data is necessary for 
high-level computations on trees.

Objects of class \code{party} basically consist of a \code{partynode}
object representing the tree structure in a recursive way and 
data. The \code{data} argument takes a \code{data.frame} which, however,
might have zero columns. Optionally, a \code{data.frame} with at least one
variable \code{(fitted)} containing the terminal node numbers of
data used for fitting the tree may be specified along with a   
\code{terms} object or any additional (currently unstructured)
information as \code{info}. Argument \code{names} defines names
for all nodes in \code{node}.

@@

\section{My first tree}

@@

In order to actually grow a tree on data, 
we have to set-up the recursion for growing a recursive 
\code{partynode} structure:
<<mytree-2, echo = TRUE>>=
growtree <- function(id = 1L, response, data, weights) {

@@

A very rough sketch of formula-based user-interface needs
to set-up the data and call \code{growtree()}:
<<mytree-3, echo = TRUE>>=
mytree <- function(formula, data, weights = NULL) {

@@

We now can fit this tree, for example to the airquality data; the
\code{print()} method provides us with a first overview on the 
resulting model
<<mytree-4, echo = TRUE>>=
aqt <- mytree(Ozone ~ Solar.R + Wind + Temp, data = airquality)
aqt
@

@@

We depict the model graphically using \code{plot()} (see Figure~\ref{plottree})
and compute predictions using
<<mytree-6, echo = TRUE>>=
predict(aqt, newdata = airquality[1:10,])
@

An interesting feature is the ability to extract subsets of trees:
<<mytree-7, echo = TRUE>>=
aqt4 <- aqt[4]
aqt4
@
which again are objects inheriting from \code{party} and thus
can be plotted easily (see Figure~\ref{subtree}).

>>>>

\author{Torsten Hothorn\\Ludwig-Maximilians-Universit\"at \\ M\"unchen
   \And Achim Zeileis\\Universit\"at Innsbruck}
\Plainauthor{Torsten Hothorn, Achim Zeileis}

@@

  The purpose of the package is to provide a toolkit with infrastructure for representing,
  summarizing, and visualizing tree-structured regression and classification models.
  Thus, the focus is not on \emph{inferring} such a tree structure from data but
  to \emph{represent} a given tree so that print/plotting and computing predictions
  can be performed in a standardized way. In particular, this unified infrastructure can be
  used for reading/coercing tree models from different sources
  (\pkg{rpart}, \pkg{RWeka}, PMML) yielding objects that share functionality for
  \fct{print}/\fct{plot}/\fct{predict} methods.
}
\Keywords{recursive partitioning, regression trees, classification trees, decision trees}

@@

\section{Motivating example and overview} \label{sec:intro}

@@

Hence, in a third step the recursive tree structure stored
in \code{pn} is coupled with the corresponding data in a \class{party}
object.
<<weather-party>>=
py <- party(pn, WeatherPlay)
py
@
And Figure~\ref{weather-plot} can easily be created by
<<weahter-plot, eval=FALSE>>=
plot(py)
@
In addition to the \class{partynode} and the \class{data.frame},
the function \fct{party} takes several further arguments 
\begin{Code}
  party(node, data, fitted = NULL, terms = NULL, ..., info = NULL)
\end{Code}
i.e., \code{fitted} values, a \code{terms} object, arbitrary additional
\code{info}, and again some further arguments comprised in \code{...}.

\subsubsection*{Further information}

More detailed technical information is provided in the subsequent sections.
Section~\ref{sec:splits}, \ref{sec:nodes}, and~\ref{sec:trees} discuss
the \fct{partysplit}, \fct{partynode}, and \fct{party} functions, respectively.
Section~\ref{sec:application} illustrates how this infrastructure can
be employed in a function that recursively \emph{infers} a tree.

\subsubsection{Design principles}

To facilitate reading of the subsequent sections, two design principles
employed in the creation of \pkg{partykit} are briefly explained.

(1)~Many helper utilities are encapsulated in functions that follow
a simple naming convention. To extract/compute some information \emph{foo}
from splits, nodes, or trees, \pkg{partykit} provides
\emph{foo}\code{_split}, \emph{foo}\code{_node}, \emph{foo}\code{_party}
functions (that are applicable to \class{partysplit}, \class{partynode},
and \class{party} objects, repectively). An example for the information
\emph{foo} might be \code{kidids}. Such functions are typically
not to be called by the end-user but potentially by package designers that
want to build functionality on top of \pkg{partykit}.

(2)~Printing and plotting relies on \emph{panel functions} that visualize
and/or format certain aspects of the resulting display, e.g., that of inner nodes,
terminal nodes, headers, footers, etc. A simple example would be printing
with a custom panel function for formatting the terminal node:
<<weather-party-print>>=
print(py,
  terminal_panel = function(node) paste(": play=", node$info, sep = ""))
@
Furthermore, arguments like \code{terminal_panel} can also take
\emph{panel-generating functions}, i.e., functions that produce a panel
function when applied to the \class{party} object.

\section{Splits} \label{sec:splits}

A split is basically a function that maps data, 
more specifically a partitioning variable, to daugther nodes. 
Objects of class \class{partysplit} are designed to represent 
such functions and are set-up by the \fct{partysplit} 
constructor:
<<partysplit-1, echo = TRUE>>=
## binary split in numeric variable `Sepal.Length'
sl5 <- partysplit(which(names(iris) == "Sepal.Length"), breaks = 5)
class(sl5)
@
The internal structure of class \class{partysplit} contains
information about the partitioning variable, the split-points,
the handling of split-points, the treatment of observations with
missing values and the daughter nodes to send observations to:
<<partysplit-2, echo = TRUE>>=
unclass(sl5)
@
Here, the split is defined in the first variable (corresponds to
\code{Sepal.Length} in data frame \code{iris}) and the splitting rule
is \code{Sepal.Length} $\le 5$:
<<partysplit-3, echo = TRUE>>=
character_split(sl5, data = iris)
@
This representation of splits is completely abstract and, most importantly,
independent of any data. Now, data comes into play when we actually
want to perform splits:
<<partysplit-4, echo = TRUE>>= 
kidids_split(sl5, data = iris)
@
For each observation in \code{iris} the split is performed and the 
number of the daughter node to send this observation to
is returned. Of course, this is a very complicated way of saying
<<partysplit-5, echo = TRUE>>=
(!with(iris, Sepal.Length <= 5)) + 1
@

@@

Let $\mu = (-\infty, \mu_1, \dots, \mu_{l - 1}, \infty)$ denote the split points
($(\mu_1, \dots, \mu_{l - 1})$ = \code{breaks}). We are interested to split according to the 
information contained in the $i$th element of $x$ ($i$ = \code{varid}).
For numeric $x_i$, the split points are also numeric. If $x_i$ is a factor at
levels $1, \dots, K$, the default split points are $\mu = (-\infty, 1, \dots, K - 1, \infty)$.

@@

The slots of \class{partysplit} objects shall be accessed by the corresponding
accessor functions.

\section{Nodes} \label{sec:nodes}

Inner and terminal nodes are represented by objects of class \class{partynode}.
Each node has a unique identifier \code{id}. A node consisting only
of such an identifier (and possibly additional information  
in \code{info}) is a terminal node:
<<partynode-1, echo = TRUE>>=
n1 <- partynode(id = 1L)
is.terminal(n1)
print(n1)
@
Inner nodes have to have a primary split \code{split} and at least two
daugther nodes. The daugther nodes are objects of class \class{partynode}
itself and thus represent the recursive nature of this data structure.
The daugther nodes are pooled in a list \code{kids}. 
In addition, a list of \class{partysplit} objects offering 
surrogate splits can be supplied; a list of \class{partysplit} objects
in slot \code{surrogates} defines such additional splits (mostly used 
for handling missing values). Note that \class{partynode} objects aren't 
connected to the actual data.

@@

A number of methods are defined for \class{partynode} objects.
\fct{is.partynode} checks if the argument is a valid \class{partynode}
object. \fct{is.terminal} is \code{TRUE} for terminal nodes
and \code{FALSE} for inner nodes. The subset methods
return the \class{partynode} object corresponding to the \code{i}th
kid:
<<partynode-5, echo = TRUE>>=
n1[2]
@

The \fct{as.partynode} and \fct{as.list} methods can be used
to convert flat list structures into recursive \class{partynode}
objects and vice versa. \fct{as.partynode} applied to
\class{partynode} objects renumbers the recursive nodes
starting with root node identifier \code{from}.

\fct{length} gives the number of kid nodes of the root node,
\fct{depth} the depth of the tree and \fct{width}
the number of terminal nodes.

\section{Trees} \label{sec:trees}

@@

Although tree structures can be represented by \class{partynode} objects,
a tree is more than a number of nodes and splits. More
information about (parts of the) corresponding data is necessary for 
high-level computations on trees.

Objects of class \class{party} basically consist of a \class{partynode}
object representing the tree structure in a recursive way and 
data. The \code{data} argument takes a \class{data.frame} which, however,
might have zero columns. Optionally, a \class{data.frame} with at least one
variable \code{(fitted)} containing the terminal node numbers of
data used for fitting the tree may be specified along with a   
\code{terms} object or any additional (currently unstructured)
information as \code{info}. Argument \code{names} defines names
for all nodes in \code{node}.

@@

\section{My first tree} \label{sec:application}

@@

In order to actually grow a tree on data, 
we have to set-up the recursion for growing a recursive 
\class{partynode} structure:
<<mytree-2, echo = TRUE>>=
growtree <- function(id = 1L, response, data, weights) {

@@

A very rough sketch of formula-based user-interface needs
to set-up the data and call \fct{growtree}:
<<mytree-3, echo = TRUE>>=
mytree <- function(formula, data, weights = NULL) {

@@

We now can fit this tree, for example to the airquality data; the
\fct{print} method provides us with a first overview on the 
resulting model
<<mytree-4, echo = TRUE>>=
aqt <- mytree(Ozone ~ Solar.R + Wind + Temp, data = airquality)
aqt
@

@@

We depict the model graphically using \fct{plot} (see Figure~\ref{plottree})
and compute predictions using
<<mytree-6, echo = TRUE>>=
predict(aqt, newdata = airquality[1:10,])
@

An interesting feature is the ability to extract subsets of trees:
<<mytree-7, echo = TRUE>>=
aqt4 <- aqt[4]
aqt4
@
which again are objects inheriting from \class{party} and thus
can be plotted easily (see Figure~\ref{subtree}).
